---
title: "LC-MS ^13^C Lactate Measurements"
date-modified: today
format: html
excecute:
  message: false
---

```{r}
#| label: setup

devtools::load_all()
suppressPackageStartupMessages({
  library(tidyverse)
  library(wmo)
  library(targets)
})

here::i_am("analysis/glc6-lactate.qmd")
here::here()

theme_set(theme_wmo())
```

```{r}
#| label: data

withr::with_dir(here::here(), {
  se_raw <- tar_read(fluxes_glc6_se)
  df_drift <- tar_read(fluxes_glc6_drift)
})
```

## Overview

To validate the results of enzymatic lactate assays suggesting decreased glycolysis in hypoxia, cells were treated with [U-^13^C~6~]-glucose (8 mM) in MCDB131 medium supplemented with glutamine (1 mM). DNA samples were collected for PicoGreen assay and medium samples were collected for LC-MS-based analysis of [u-^13^C~3~]-lactate concentration. Medium samples (10 μL) were combined with D~8~-valine (10 μL of 100 100 μM) and extracted with 180 μL MeOH pre-cooled to -80 °C. Samples were spun and the supernatants were analyzed by LC-MS using both full scan and selected ion monitoring with negative polarity. Quality control samples were interspersed across all of the samples to assess for instrument drift. 

# Analysis

```{r}
#| label: raw-data
# df <-
#   read_ms_excel(here::here(path)) |> 
#   dplyr::filter(!str_detect(.data$id, "B")) |> 
#   tidyr::separate(
#     .data$sample, 
#     c("type", "number"), 
#     sep = "(?<=[[:alpha:]])(?=[[:digit:]])", 
#     fill = "left", 
#     convert = TRUE
#   ) |>
#   dplyr::mutate(
#     conc = ifelse(is.na(.data$type), .data$number, NA_character_), 
#     date = dplyr::case_when(
#       .data$type == "A" ~ "2022-11-12", 
#       .data$type == "B" ~ "2022-11-17", 
#       .data$type == "C" ~ "2022-11-22"
#     ), 
#     number = ifelse(!is.na(conc), NA, .data$number), 
#     batch = "b", 
#     cell_type = "lf", 
#     experiment = "substrate", 
#     type = dplyr::case_when(
#       .data$type == "QC" ~ "qc", 
#       is.na(.data$type) ~ "std", 
#       TRUE ~ "sample"
#     ), 
#     detector = "lcms", 
#     run = "a"
#   ) |> 
#   dplyr::select(
#     "metabolite", 
#     "detector", 
#     "mz", 
#     "rt", 
#     "cell_type",
#     "experiment", 
#     "batch", 
#     "date", 
#     "run", 
#     "id", 
#     "type", 
#     "number", 
#     "conc", 
#     "area"
#   )
```

## Drift

```{r}
#| label: drift-plot

df_raw <- 
  se_raw |> 
  se_to_tbl(rownames = "metabolite") |> 
  filter(type == "qc") |> 
  group_by(metabolite) |> 
  mutate(norm = area / mean(area))

ggplot(df_raw) +
  aes(
    x = number, 
    y = norm, 
    color = metabolite
  ) +
  geom_line() + 
  labs(
    x = "Run order", 
    y = "Normalized peak area", 
    color = "Metabolite", 
    title = "QC Drift"
  )
```

The QC peak areas suggest that correcting for experimental drift may be helpful over the time course of the experiment.

```{r}
#| label: qc-variance

ggplot(df_raw) +
  aes(
    x = metabolite, 
    y = norm, 
    color = metabolite
  ) +
  geom_boxplot(show.legend = FALSE) +
  geom_jitter(
    width = 0.1, 
    show.legend = FALSE
  ) +
  labs(
    x = "Metabolite", 
    y = "Normalized peak area", 
    title = "QC Variance"
  )
```

Regardless of drift, the QC peak areas are quite similar across the experimental time course.

```{r}
#| label: drift-corr-sanity-check

df_drift |> 
  filter(type == "qc") |> 
  group_by(metabolite) |> 
  mutate(norm = area / mean(area)) |> 
  ggplot() +
  aes(
    x = number, 
    y = norm, 
    color = metabolite
  ) +
  geom_line() + 
  labs(
    x = "Run order", 
    y = "Normalized peak area", 
    color = "Metabolite", 
    title = "Drift Corrected"
  )
```

## Standard Curves

```{r}
#| label: standard-curves

df_drift |> 
  dplyr::filter(type == "std" & metabolite == "LAC M0") |> 
  dplyr::mutate(conc = as.numeric(conc)) |> 
  ggplot() +
  aes(
    x = conc, 
    y = area
  ) +
  geom_point()
```

## Samples

```{r}
#| label: samples

pheno <- 
  tibble::tibble(
    number = 1:24, 
    oxygen = rep(c("21%", "0.5%"), times = 4, each = 3), 
    time = rep(c(0, 24, 48, 72), each = 6), 
    rep = rep(c("x", "y", "z"), 8)
  )

df_drift |> 
  filter(type %in% letters[1:3]) |> 
  filter(metabolite == "LAC M0") |> 
  left_join(pheno, by = "number") |> 
  group_by(type, oxygen, time) |> 
  wmo::remove_nested_outliers(area, remove = TRUE) |> 
  summarize(area = mean(area)) |> 
  ggplot() +
  aes(
    x = time, 
    y = area, 
    color = oxygen
  ) +
  geom_point() +
  stat_summary(
    geom = "line"
  )
```

These plots suggest that, for the 100 μL extraction buffer, the maximum cell count is ~ 250,000. We will exclude the 300,000 value prior to calculating the slopes of the curves. For the PASMC samples, the 400,000 cell sample is quite high compared to the other samples. I will also remove these values prior to calculation of the slopes.

