---
title: "LC-MS ^13^C Lactate Measurements"
date-modified: today
format: html
excecute:
  message: false
---

```{r}
#| label: setup
#| output: false

devtools::load_all()
suppressPackageStartupMessages({
  library(tidyverse)
  library(wmo)
  library(targets)
})

here::i_am("analysis/glc6-lactate.qmd")

theme_set(theme_wmo())
```

```{r}
#| label: data

withr::with_dir(here::here(), {
  se_raw <- tar_read(fluxes_glc6_se)
  df_drift <- tar_read(fluxes_glc6_drift)
  std <- tar_read(conc_std)
  fluxes <- tar_read(fluxes)
})
```

## Overview

To validate the results of enzymatic lactate assays suggesting decreased glycolysis in hypoxia, cells were treated with [U-^13^C~6~]-glucose (8 mM) in MCDB131 medium supplemented with glutamine (1 mM). DNA samples were collected for PicoGreen assay and medium samples were collected for LC-MS-based analysis of [u-^13^C~3~]-lactate concentration. Medium samples (10 μL) were combined with D~8~-valine (10 μL of 100 100 μM) and extracted with 180 μL MeOH pre-cooled to -80 °C. Samples were spun and the supernatants were analyzed by LC-MS using both full scan and selected ion monitoring with negative polarity. Quality control samples were interspersed across all of the samples to assess for instrument drift. 

# Analysis

## Drift Correction

```{r}
#| label: drift-plot

df_raw <- 
  se_raw |> 
  se_to_tbl(rownames = "metabolite") |> 
  filter(type == "qc") |> 
  group_by(metabolite) |> 
  mutate(norm = area / mean(area))

ggplot(df_raw) +
  aes(
    x = number, 
    y = norm, 
    color = metabolite
  ) +
  geom_line() + 
  labs(
    x = "Run order", 
    y = "Normalized peak area", 
    color = "Metabolite", 
    title = "QC Drift"
  )
```

The QC peak areas suggest that correcting for experimental drift may be helpful over the time course of the experiment.

```{r}
#| label: qc-variance

ggplot(df_raw) +
  aes(
    x = metabolite, 
    y = norm, 
    color = metabolite
  ) +
  geom_boxplot(show.legend = FALSE) +
  geom_jitter(
    width = 0.1, 
    show.legend = FALSE
  ) +
  labs(
    x = "Metabolite", 
    y = "Normalized peak area", 
    title = "QC Variance"
  )
```

Regardless of drift, the QC peak areas are quite similar across the experimental time course.

```{r}
#| label: drift-corr-sanity-check

df_drift |> 
  filter(type == "qc") |> 
  group_by(metabolite) |> 
  mutate(norm = area / mean(area)) |> 
  ggplot() +
  aes(
    x = number, 
    y = norm, 
    color = metabolite
  ) +
  geom_line() + 
  labs(
    x = "Run order", 
    y = "Normalized peak area", 
    color = "Metabolite", 
    title = "Drift Corrected"
  )
```

## Standard Curves

```{r}
#| label: standard-curves
#| results: hide

std |> 
  filter(experiment == "substrate" & metabolite == "lactate") |> 
  pull(plots)
```

Lactate peak areas were normalized to the internal standard peak area to generate a standard curve. Sample concentrations were interpolated after correcting the M3 peak areas for quadrupole bias prior to flux calculations.

## Samples

```{r}
#| label: samples

fluxes |> 
  filter(experiment == "substrate") |> 
  ggplot() +
  aes(
    x = oxygen, 
    y = flux, 
    fill = oxygen
  ) +
   stat_summary(
    geom = "col", 
    fun = "mean", 
    show.legend = FALSE
  ) +
  geom_point(
    color = "white", 
    pch = 21, 
    show.legend = FALSE
  ) + 
  stat_summary(
    geom = "errorbar", 
    width = 0.2, 
    fun.data = "mean_se", 
    linewidth = 0.25
  ) +
  labs(
    x = "Oxygen", 
    y = "Flux (fmol/cell/h)"
  )
```

These data are consistent with the results obtained from enzymatic measurements of extracellular lactate demonstrating decreased lactate efflux from glycolysis in hypoxia compared to normoxia.
