---
title: "Q Bias Correction Factors"
date-modified: today
format: html
excecute:
  message: false
---

```{r}
#| label: setup
#| output: false

devtools::load_all()
suppressPackageStartupMessages({
  library(tidyverse)
  library(wmo)
  library(targets)
})

here::i_am("analysis/qbias.qmd")
here::here()

theme_set(theme_wmo())
```

```{r}
#| label: data

withr::with_dir(here::here(), {
  qbias_ratios <- tar_read(qbias_ratios)
  correction_factors <- tar_read(correction_factors)
})
```

# Overview

A selected ion monitoring method was utilized to increase sensitivity for isotope detection. Others have noted quadrupole bias on the Q Exactive using this detection method (*i.e.*, the sensitivity for each isotope differs relative to its position within the m/*z* window). To account for this, naturally labeled standards were analyzed using methods where the m/*z* window is shifted -1.003355 units, the ^13^C mass defect. 

# Analysis

The M1/M0 ratio was calculated for each frame shift. The resulting data were fit with a third-degree polynomial function. 

```{r}
#| label: functions

make_plots <- function(data, batch, metabolite) {
  ggplot(data) +
    aes(x = window, y = ratio) +
    geom_smooth(
      method = MASS::rlm, 
      formula = y ~ poly(x, 3),
      method.args = list(maxit = 1000), 
      color = "blue", 
      se = FALSE
    ) +
    geom_point(
      size = 2, 
      alpha = 0.5, 
      color = "black"
    ) +
    geom_point(
      size = 5, 
      stat = "summary", 
      fun.data = "mean_se", 
      color = "blue"
    ) +
    labs(
      title = str_c(metabolite, batch, sep = " "), 
      x = "Window", 
      y = "Ratio"
    ) 
}
```

```{r}
#| label: isotope-ratios
#| layout-ncol: 4
#| results: hide

qbias_ratios |> 
  nest() |> 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) |> 
  pull(plots)
```

Q bias correction linearizes the isotope ratios across the measurement window. 

```{r}
#| label: sanity-check-correction
#| layout-ncol: 4
#| results: hide

qbias_ratios |> 
  select(-ratio) |> 
  pivot_longer(c(M0, M1), names_to = "isotope", values_to = "value") |> 
  mutate(
    M = str_extract(isotope, "\\d") |> as.numeric(), 
    M = M + window, 
    iso = str_c("M", M)
  ) |> 
  left_join(correction_factors, by = c("metabolite", "batch", "iso" = "M")) |> 
  mutate(value = value * cf) |> 
  select(-c(cf, iso, M)) |> 
  pivot_wider(names_from = isotope, values_from = value) |> 
  mutate(ratio = M1 / M0) |> 
  group_by(metabolite, batch) |> 
  nest() |> 
  mutate(plots = pmap(list(data, batch, metabolite), make_plots)) |> 
  pull(plots)
```

